---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

First we'll load the activity.csv from the downloaded dataset and the transform the 'Date' column into the date format.
```{r Load}
activity<-read.csv("activity.csv")
activity$date<-as.Date(activity$date, format="%Y-%m-%d")
```


Then we'll sum the steps aggregated by date
```{r Steps}
stepsbyday<-aggregate(activity$steps,by=list(Date=as.Date(activity$date)),FUN=sum)
```


## What is mean total number of steps taken per day?

We'll look at a histogram of the total steps in a day.

```{r Histogram}
library(ggplot2)
qplot(stepsbyday$x, data=stepsbyday, geom="histogram", xlab="Total Steps in a Day", ylab="Number of Days", main="Histogram of Number of Steps Per Day")
```

## What is the average daily activity pattern?

Then we'll look at the mean and median steps by day.

```{r stats }
meansteps<-mean(stepsbyday$x,na.rm = TRUE)
mediansteps<-median(stepsbyday$x,na.rm = TRUE)
meansteps
mediansteps
```

Now we'll take the mean of the steps aggregated by the interval and insert into a new frame.

```{r byinterval}
stepsbyinterval<-aggregate(activity$steps,by=list(interval=activity$interval),FUN=mean, na.rm=TRUE)
```

Then plot the mean of the steps by interval

```{r intervalplot}
qplot(x=interval, y=x, data=stepsbyinterval, geom="line",ylab="Average Steps", main="Average Steps by interval")
```

Max steps.

```{r maxsteps}
maxsteps<-stepsbyinterval[stepsbyinterval$x==max(stepsbyinterval$x),]
maxsteps
```

## Imputing missing values

We'll see how many values we're missing

```{r missingdays}
nummissing<-sum(is.na(activity$steps))
nummissing
```

We'll fill in the missing values with the mean steps of that interval

```{r fillmissing}
myfxn<-function(steps, interval)
      {if(is.na(steps)==TRUE)
          {stepsbyinterval[stepsbyinterval$interval==interval,2]}
      else{steps}}
activityfilled<-activity
activityfilled$steps<-mapply(myfxn, activity[,1],activity[,3])
```

No we'll re-plot the histogram with the filled in data. This moves the histogram to the right and makes it taller.

```{r FilledHistogram}
filledstepsbyday<-aggregate(activityfilled$steps,by=list(Date=as.Date(activityfilled$date)),FUN=sum)
qplot(filledstepsbyday$x, data=filledstepsbyday, geom="histogram", xlab="Total Steps in a Day", ylab="Number of Days", main="Histogram of Number of Steps Per Day, NA filled with average", bins=20)
```

Finding the mean and median steps of the filled data

```{r filledstats}
meanstepsfilled<-mean(filledstepsbyday$x,na.rm = TRUE)
medianstepsfilled<-median(filledstepsbyday$x,na.rm = TRUE)
meanstepsfilled
medianstepsfilled
```

## Are there differences in activity patterns between weekdays and weekends?

Creating a factor of "Weekday" and "Weekend"

```{r weekend}
activityfilled$Weekend<-factor(weekdays(as.Date(activityfilled$date), abbreviate=FALSE), levels=c("Monday", "Tuesday", "Wednesday","Thursday","Friday","Saturday","Sunday"),labels = c("Weekday","Weekday","Weekday","Weekday","Weekday","Weekend", "Weekend"))
```

Aggregating on interval and whether it's a weekday or weekend

```{r weekendaggregate}
stepsbyintervalandweekend<-aggregate(activityfilled$steps,by=list(interval=activityfilled$interval, weekend=activityfilled$Weekend),FUN=mean, na.rm=TRUE)
```

Make a new line plot of steps by interval broken out into weekend and weekday

```{r weekendplot}
qplot(x=interval, y=x, data=stepsbyintervalandweekend, facets=stepsbyintervalandweekend$weekend~., geom="line", xlab="Interval",ylab="Number of Steps", main="Mean Steps Per Interval by Weekend and Weekday")

```
